---
title: "tutorials"
output: html_document
---

## Atomic Vectors

### Task 1

Create a vector x containing elements 1 through 10. Now, apply the dim() function to x.

```{r}

vec <- 1:10
dim(vec)

```

This returns NULL because dim is specifically designed for objects with more than one dimension, such as matrices or arrays.

```{r}
x <- 1:10 ; mean(x); x[2] <- NA ; mean(x)  
mean(x, na.rm = TRUE)

```

## Rectangular Data

### Task 1

Create an atomic vector x by concatenating integers 1 to 8. Use the following code, or come with your own solution.

```{r}
x <- 1:8; x

```

```{r}
attributes(x)
rownames(x)
colnames(x)
```

### Task 5

```{r}
l <- list()

for (i in 1:3) {
  l[[i]] <- rnorm(n = 3, mean = i, sd = 0.5)
}

m <- matrix(nrow = 0, ncol = 3, byrow = TRUE)

for (vec in l) {
  m <- rbind(m, vec)
}

m
```

### Task 6

```{r}

test <- c(5,6)

typeof(test)

typeof(m)
class(m)
```

### Task 7

```{r}
dim(m) <- NULL
m
class(m)
typeof(m)
```

### Task 8

```{r}

a <- array(m, dim = c(2, 3))
a

class(a)
```

```{r}

set.seed(123)

# Build an empty matrix with 3 columns and no rows. 
m <- matrix(nrow = 0, ncol = 3, byrow = TRUE) 

# Fill an empty matrix with randomly generated triplets.
for (i in 1:3) m <- rbind(m, rnorm(n = 3, mean = i, sd = 0.5))

# Name rows and columns.
colnames(m) <- c("Alpha", "Beta", "Gamma")
rownames(m) <- c("Alpha", "Beta", "Gamma") ; m

class(m)
typeof(m)

dim(m) <- NULL

class(m)

m_as_array <- array(m, dim = c(2,3))

class(m_as_array)
```

```{r}

test_vec <- c(1,2,3)

colnames(test_vec) <- c("bullshit", "piča", "kokot")

length(test_vec)
names(test_vec)
```

## Subsetting

```{r}

m[1,(2:3)]

```

```{r}
List1 <- list(list(1, 2), c(3, 4))
List1
List2 <- c(list(1, 2), c(3, 4))
class(List2)

```

unlist(

## 2D Data

## Recursive Vectors & Data Frames

#### Task 3

Here List1 contains recursive lists and List2 is created because atomic vectors cannot hold other collections.

```{r}
List1 <- list(list(1, 2), c(3, 4))
List1
class(List1)
List2 <- c(list(1, 2), c(3, 4))
class(List2)

```

#### Task 4

In R, both **atomic vectors** and **lists** are technically "vectors," but they serve very different purposes based on how they handle data types and memory.

##### Similarities

-   Both are vectors

-   We can use [] for indexing

-   Both can have names for their elements, as well as other attributes like dim or class

-   Both have a lenth()

##### Differences

-   Homegeneity vs heterogeneity

-   Vectors are flat (a simple sequence of values) and lists are recursive

-   Vector's coercion forces all elements to the most flexible type (c(1, "a") -\> "1", "a") whilsts lists keep all elements in their original form.

##### Where c() will work and list() wont

```{r}
# Using c() - This works!
x <- c(10, 20, 30)
x + 1 
# Result: 11, 21, 31

# Using list() - This fails!
y <- list(10, 20, 30)
y + 1
# Result: Error in y + 1 : non-numeric argument to binary operator
```

#### Task 5

```{r}

unlistedList1 <- unlist(List1)
unlistedList1

unlistedList2 <- unlist(List2)
unlistedList2
```

#### Task 6

```{r}
List3 <- list(list(list(1)))
str(List3)

```

#### Task 7

The most important list application in R is a data frame. A data frame contains list of column-bound vectors which can have attributes. In contrast to a regular list, a data frame has an additional constraint: the length of each of its vectors must be the same. This gives data frames their rectangular structure and explains why they share some properties with matrices.

#### Modifying a column vs row

-   R raplaces the original column with the new one (copy-on-modify)

-   When modifying a row R needs to replace every column vector

-   This is because data frames are column-oriented structures oriented for column operations.

#### Subsetting

🔹 [ → slicing / subsetting

Returns the same type of container

Just with fewer elements

🔹 [[ → extraction

Returns the actual element inside

Removes one level of structure

```{r}

Players <- list(Offense = data.frame(Position = c('Striker', 'Left Winger'), Wage = c(2.1, 0.49)),
                Midfielders = data.frame(Position = c('Center', 'Right Midfielder'), Wage = c(2.18, 3.89)),
                Defense = data.frame(Position = c('Goal Keeper', 'Right Center Back'), Wage = c(0.99, 0.69)))

class(Players)


```

##### Using [ vs [[

```{r}

class(Players[1])
Players[1]

class(Players[[1]])
Players[[1]]
```

##### \$ shorthand

An alternative way of doing the same thing is using the dollar sign operator $, which is a shorthand operator: x$y is roughly equivalent to x[["y"]]. It’s often used to access variables in a data frame, which has named columns.

The one important difference between \$ and [[]] is that \$ does (left-to-right) partial matching, as shown below:

```{r}
Players$Off # Off is partially matched to Offense, which is returned.
```

## Loading data into R

### Libs

```{r}
library(tidyverse)
library(readr)

```

### Task 1

```{r}
# pattern = "player" tells list.files() to return only filenames that contain "player" anywhere in the name.
list.files("./data/original", include.dirs = FALSE, pattern = "player")
```

### Loading data

```{r}
data_tibble <- read_csv("./data/original/players_22.csv")
data_df <- read.csv("./data/original/players_22.csv")
str(data_tibble)
```

### Task 2 - read.csv vs read_csv

| Feature | `read.csv()` (base R) | `read_csv()` (readr) |
|------------------|------------------------|------------------------------|
| **Package** | Base R (no install needed) | readr (tidyverse) |
| **Speed** | Slower (especially large files) | Much faster |
| **Return type** | `data.frame` | `tibble` (modern data frame) |
| **Strings handling** | Converts to factors by default *(older R)* | Always keeps as character |
| **Missing values** | Recognizes `NA` and empty strings (`""`) | Recognizes `NA`, `""`, and more (e.g. `"N/A"`, `"NULL"`) |
| **Type inference** | Less consistent, sometimes guesses poorly | Strong, consistent type guessing |
| **Column types** | Often converts text → factor | Keeps text as character unless specified |
| **Printing** | Prints entire data frame | Prints preview (cleaner) |
| **Column parsing feedback** | No messages | Shows parsing info + problems |
| **Error handling** | Silent failures possible | Explicit warnings and diagnostics |

### Task 3

```{r}
getwd()
df <- read_csv("./data/original/players_22.csv", col_names = FALSE)
colnames(df)
```

### Task 4

Here t() transposes data frame and changes colnames -\> rownames

```{r}

transposed <- t(data_df)

colnames(transposed)
rownames(transposed)
```

### Task 5 - Combining data franes

`rbind()` stacks data **row-wise** (one dataset under the other)

This works **only if both datasets have the same columns** (which FIFA datasets typically do)

```{r}

players_21 <- read_csv("./data/original/players_21.csv")
players_22 <- read_csv("./data/original/players_22.csv")

players_all <- rbind(players_21, players_22)

str(players_all)
```

### Task 6

```{r}

players_22_selected <- read_csv("./data/original/players_22.csv", col_select = c(short_name, contains("skill"), contains("id")))
```

### Task 7 - NA filtering

```{r}

players_22_selected <- read_csv("./data/original/players_22.csv", na = c('', "NA"))

```

### Slice

```{r}

slice(data_tibble, c(1:3), overall > 90)

```

## Functions

### Libs

```{r}



```

### Task 1

in R:

### 1. Functions are **first-class objects**

They don’t inherently store their “name”.

test_func is just a pointer to the function and there can be more pointers pointing to the same function object.

```{r}


test_func <- function(x, y) {
  return(x + y)
}

match.fun("test_func")
y <- c(5)
primitive <- 5
```

### Task 2

```{r}

is.function(y)
is.primitive(y)

```

```{r}

is.function(y)
is.primitive(primitive)
```

### Default Arguments

```{r}

greet <- function(name, greeting = "Fuck you"){
  return(paste(greeting, name))
}

message1 <- greet("alice")
print(message1)

```

### Task 3

R uses **lexical scoping**, so when you redefine a function name, you **mask (override)** the original function in the current environment.

```{r}

mean <- function(x){
  return("pica")
}

mean(1:5)

#You can still access it explicitly:
base::mean(1:5)
```

### Task 4

```{r}

sum(c(1, 2, 3))
mean(c(1, 2, 3))

sum(1, 2, 3, na.omit = TRUE) 
mean(1, 2, 3, na.omit = TRUE)
```

### Task 5

```{r}
f <- function(abcdef, bcde1, bcde2) {   
  list(a = abcdef, b1 = bcde1, b2 = bcde2) 
}
str(f(1, 2, 3)) 
str(f(2, 3, abcdef = 1)) 
str(f(2, 3, a = 1))

```

### Task 6

```{r}

data_frame_setup <- function(...){
  return(
    data.frame(
      ...
    )
  )
}

a <- 1:10 
b <- 2:11  
c <- 3:12

data_frame <- data_frame_setup(a, b, c)

str(data_frame)

data_frame$a
```

### Task 7

No, anonymous functions can taky any number of arguments.

### Task 8

Here we nest an anonymous function inside named function apply_multiplier and pass it to the sapply() function.

```{r}
apply_multiplier <- function(x, factor) {
  sapply(x, function(y) y * factor)
}

apply_multiplier(1:5, 4)
```

## Lexical Scoping

## A Grammar of Data Manipulation

### Libs

```{r}

library(tidyverse)
library(readr)
library(dplyr)

```

### Task 1

**Longitudinal**: same players are analyzed across a certain period of time (2019-2022)

**cross-sectional**: each file only collects data from a single point in time

### Task 2

```{r}

players_22 <- read_csv("./data/original/players_22.csv")

```

```{r}
# files <- paste0("players_", 15:22, ".csv")
# 
# # gsub(pattern, replacement, x)
# # sub - only first match, gsub all
# # pattern → regex pattern to search for
# # replacement → string to replace it with
# # x → character vector
# 
# # "players_|\\.csv" → the | means “or” in regex
# 
# # \\. escaped dot, because . is special in regex
# 
# # Replaces "players_" or ".csv" with "" (nothing)
# all_players <- bind_rows(
#   lapply(files, function(f) {
#     read_csv(paste0("./data/original/", f)) %>%
#       mutate(season = as.integer(gsub("players_|\\.csv", "", f)))
#   })
# )
```

### Task 3

```{r}

nrow(players_22)
ncol(players_22)

table(sapply(players_22, class))

```

### Task 4

```{r}
View(players_22)
skill_colls <- function(data_frame, player_name){
  return(
    data_frame %>% filter(short_name == player_name) %>% select(starts_with("skill"), ends_with("ing"))
  )
}


skill_colls(players_22, "L. Messi")
```

### Task 5

```{r}

data %>%
  mutate(
    across(starts_with("skill_"),
           ~ .x - median(.x, na.rm = TRUE),
           .names = "{.col}_mednorm"
    )
  )
```

### Task 7

```{r}

data_normalized <- data %>%
  select(short_name, dob, skill_dribbling) %>%
  filter(short_name %in% c("R. Lewandowski", "Cristiano Ronaldo", "Neymar Jr")) %>%
  mutate(skill_dribbling = skill_dribbling - median(skill_dribbling, na.rm = TRUE))
```

### Task 8

#### Without replacement

-   Once picked the element cannot be picked again

```{r}
sampled_rows <- players_22 %>%
  sample(n = 1000)
```

#### With replacement

-   Can be picked multiple times

```{r}
samples_rows <- players_22 %>%
  sample(n=1000, replace = TRUE)
```

## Data Summaries

### Summary

```{r}

data %>% 
  select(short_name, dob, skill_dribbling, league_level) %>%
  summarise(mean_dribbling = mean(skill_dribbling),
            min_dribbling = min(skill_dribbling),
            max_dribbling = max(skill_dribbling),
            number_of_records = n())
```

### Task 1

```{r}
players_summary <- players_22 %>%
  select(where(is.numeric)) %>%
  summarise(
    across(
      starts_with("skill_"),
      list(
        mean = ~ mean(.x, na.rm = TRUE),
        min  = ~ min(.x, na.rm = TRUE),
        max  = ~ max(.x, na.rm = TRUE)
      ),
      .names = "{.col}_{.fn}"
    ),
    number_of_records = n()
  )
players_summary

```

### Task 5

```{r}

data_sorted <- data[order(data$overall), ]
```

## Data Tidying

### Task 1

```{r}
setwd("C:/Users/fmojt/DataScience/OZNAL_practicums/data/original")

```
